Evaluation of Diagnostic Concordance Between Algorithms for Parkinson’s Disease Dementia

### 2025-10-15 _manuscript/manuscript.qmd
Manuscript changes to incorporate all reviewers' comments.

According to a recent meta-analysis, approximately one-quarter of PD patients is likely to be diagnosed with PDD [@sousa2022]. However, reported PDD rate estimates vary widely, ranging from 14% up to 55%, depending on the methodological criteria employed [@sousa2022]. Moreover, factors such as patients' sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline and PDD [@oh2016; @rana2012].

@@

Currently, efforts are focused on refining this PDD diagnostic framework. A recent call for a change pinpoints limitations regarding the original criteria and suggest various updates to enhance their utility [@kulisevsky2024]. Proposed suggestions include replacement of the Mini Mental State Examination (MMSE) by the Montreal Cognitive Assessment (MoCA), which is more sensitive to PD specific cognitive impairment; expansion of instrumental activities of daily living (IADL) evaluation; inclusion of language assessment; recognition of anxiety as one **of** the neuropsychiatric symptoms relevant in PDD; and integration of biomarkers.

@@

Cognitive performance was evaluated at both Level I and Level II according to the standard MDS battery for Parkinson’s Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Cognitive performance at Level I was assessed by **the MMSE [@stepankova2014; @folstein1975] and the MoCA** [@kopecek2016; @nasreddine2005]. The comprehensive neuropsychological assessment at Level II evaluated five cognitive domains through specific tests: attention and working memory assessed by Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997], executive function by Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021], language by the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983], memory by the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall, visuospatial function assessed by the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

@@

In this study, we applied three distinct sets of diagnostic algorithms for probable PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of algorithms was based on the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocates for more sensitive cognitive domain assessments in the context of PD. This updated approach incorporated specific items from the MoCA. The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to measure global cognitive performance using a reduced testing protocol that omits items providing redundant information. **The sMoCA has been validated in Czech PD cohort [@bezdicek2020] and shown to be sensitive to cognitive deficits while lowering patient burden [@roalf2017a]. We included the sMoCA in our study for its clinical utility in pre-surgical settings, where time restrictions and patients’ fatigue often limit the feasibility of longer assessments. Moreover, the Czech validation study reported comparable diagnostic accuracy between MoCA (AUC = 0.815) and sMoCA (AUC = 0.796) for distinguishing PD-MCI from PD-NC, supporting the sMoCA as a suitable and efficient alternative.**

@@

Following the framework proposed @lundberg2021, in this study we explicitly connect our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands to statistical estimates. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. A full description of the study’s estimands and their relation to our research objectives is presented in the Appendix (see @tbl-estimands).

@@

@@

**Total of** `r n_total` patients **were considered for the study, out of which `r excl_txt`**. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. **Descriptive statistics for neuropsychiatric symptoms indicated within average level of depressive and anxiety symptoms in our cohort, with the average BDI-II of `r bdi`, average STAI X1 of `r staix1`, and average STAI X2 of `r staix2`. However, according to the psychiatric assessment, none of the patients with probable PDD was suffering from the major depressive disorder, delirium or other neuropsychiatric abnormalities that would excluded the diagnosis.** Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

```{r}
#| label: "fig-tree"
#| fig-cap: "A dendrogram representing algorithms creation."

make_dendrogram(rates$table)
```

@@

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **The number of complete pairwise cases ranged from `r obs_pairs[1, "N"]` (`r min_miss` missing) to `r obs_pairs[nrow(obs_pairs), "N"]` (full sample). Most comparisons (`r perc_above_k`%) were based off of `r K` or more observations.**[^1]

@@

table_levelII_approximations(conc$table, algos, "Accuracy_raw")
```

@@

Our overall PDD rates were consistently lower than previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigation found even higher rate, reaching up to 30% [@aarsland2005]. A recent meta-analysis synthesizing global data placed the expected PDD rate in PD at 26.30% [@sousa2022]. Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in diagnostic criteria, methodology and sample characteristics. Specifically, our sample was younger compared to other PD cohorts and age was repeatedly shown to be a strong predictor of PDD across studies [@rana2012; @sousa2022].

@@

The advantage of MoCA-based algorithm was evident in combination with using the FAQ total score, reaching accuracy rates of 94–99%. Furthermore, in the algorithm using sMoCA, the raw accuracy was moderate, however, the balanced accuracy (i.e. combined sensitivity and specificity) was high. These findings imply that simplified tools such as sMoCA may be a promising tool for heightening efficiency of the assessment while maintaining diagnostic accuracy. **This underscores the clinical utility of sMoCA screening tool. Although raw accuracy was moderate, balanced accuracy results ensure avoidance of bias from class imbalance. From a practical point of view, it suggests that even brief screening tools can achieve reliable classification, which means an advantage for routine clinical settings and time-restricted assessments.**

@@

**Modern alternatives, such as multiple imputation techniques, have been shown to produce less biased estimates than case deletion strategies in analyses based on confusion matrices [@karakaya2014], whereas the evidence supporting their superiority in the estimation of Cohen’s** $\kappa$ remains limited [@deraadt2019]. These modern approaches also require careful specification of the imputation model and the underlying causal mechanism of missingness to ensure appropriate covariate selection [@long2011; @bianco2023]. For the sake of parsimony, we did not employ advanced missing-data techniques but instead explicitly described the observed missingness patterns. Consequently, our findings should be regarded as exploratory and primarily serve as a basis for hypothesis generation in future, more controlled studies.

@@

**Another limitation concerns the use of the FAQ questionnaire for IADL assessment. The FAQ is a subjective and informant-reliant measure, which may contain biased data. Moreover, its content can vary across sociocultural contexts, limiting cross-cultural transferability. For example, activities such as financial management, cooking or driving, are not uniformly performed across societies. Consequently, both the FAQ scores and diagnostic thresholds used to indicate IADL impairment may not be directly transferable between cultural settings [@odonald2025]. These factors may influence both the sensitivity and ecological validity of the functional criteria used [@bezdicek2011; @bezdicek2016b].**

**Furthermore, the FAQ scores may correlate with neuropsychological results of specific cognitive domains, particularly executive functions or memory [@vanrentergem2020; @moheb2017]. This overlap could artificially increase concordance between cognitive and functional capacity criteria. To address these concerns in future research, we suggest using a PD-specific questionnaire, such as the Penn Parkinson's Daily Activities Questionnaire-15 [@brennan2016] or generally more objective methods, preferably performance-based ones. Our results underscore the importance of IADL measurement for the PDD diagnosis. Therefore, we recommend using more reliable tools with high ecological validity.** For future research, we suggest the use of more objective, modernized tools for IADl assessment, such as performance-based methods [@schmitter-edgecombe2020] or questionnaire adaptations including items regarding gadgets use and digital literacy [@postema2024]. Our results emphasize the importance of IADL deficit for the PDD diagnosis, therefore, we suggest exploring more reliable tools with high ecological validity.

Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. Future studies would benefit from focusing on these aspects of PDD.

@@

Future studies should aim to replicate our results on larger and different cohorts **with more heterogenous sample**. To make this process easier, the code used to generate our results is publicly available and easily applicable to similarly structured data.

@@

In this study, we follow the framework proposed by @lundberg2021 for specifying targets of inference (i.e., the estimands) in qunatitative sciences to increase transparency and connect statistical evidence to relevant theory. @tbl-estimands contains verbal description of the components relating to each of our proclaimed research objectives and map them to the population quantity of interest (the theoretical estimand), data-dependent quantity that could be estimated (the empirical estimand) and quantities that are reported in the study (statistical estimates).

@@

```{r}
#| label: "fig-rates"
#| fig-cap: "Summary of the estimates of probable PDD rate."
#| apa-note: "Vertical lines represent estimates arrived at by using sMoCA (dotted) or Level II (dashed) with FAQ item 9 (orange) or FAQ total score (blue) as criteria for probable PDD."

@@

table_levelII_approximations(conc$table, algos, "Balanced Accuracy")
```

>>>>

According to a recent meta-analysis, approximately one-quarter of PD patients are likely to be diagnosed with PDD [@sousa2022]. However, reported PDD rate estimates vary widely, ranging from 14% up to 55%, depending on the methodological criteria employed [@sousa2022]. Moreover, factors such as patients' sex [@cereda2016], age and disease duration appear to modulate the risk of cognitive decline and PDD [@oh2016; @rana2012].

@@

Currently, efforts are focused on refining this PDD diagnostic framework. A recent call for a change pinpoints limitations regarding the original criteria and suggests various updates to enhance their utility [@kulisevsky2024]. Proposed suggestions include replacement of the Mini Mental State Examination (MMSE) by the Montreal Cognitive Assessment (MoCA), which is more sensitive to PD specific cognitive impairment; expansion of instrumental activities of daily living (IADL) evaluation; inclusion of language assessment; recognition of anxiety as one **of** the neuropsychiatric symptoms relevant in PDD; and integration of biomarkers.

@@

Cognitive performance was evaluated at both Level I and Level II according to the standard MDS battery for Parkinson’s Disease Mild Cognitive Impairment (PD-MCI) [@litvan2012, @bezdicek2017]. Cognitive performance at Level I was assessed by the MMSE [@stepankova2014; @folstein1975] and the MoCA [@kopecek2016; @nasreddine2005]. The comprehensive neuropsychological assessment at Level II evaluated five cognitive domains through specific tests: attention and working memory assessed by Trail Making Test Part A (TMT-A) [@bezdicek2012; @reitan2004], and WAIS Digit Span Backward (WAIS DSB) [@wechsler1997], executive function by Categorical Verbal Fluency (CF) [@nikolai2015], and subtest from the Prague Stroop Test – Colors (PST-C) [@bezdicek2021], language by the WAIS Similarities subtest [@wechsler1997], and the Boston Naming Test (BNT-60) [@zemanova2016; @kaplan1983], memory by the Rey Auditory Verbal Learning Test (RAVLT) [@frydrychova2018; @bezdicek2013; @rey1964] delayed recall, and the Brief Visual Memory Test–Revised (BVMTR) [@havlik2020; @benedict1997] delayed recall, or WAIS Family Pictures subtest [@wechsler1997] delayed recall, visuospatial function assessed by the Judgment of Line Orientation Test (JoLO) [@benton1983], and Clock Drawing Test (CLOX) [@royall1998].

@@

In this study, we applied three distinct sets of diagnostic algorithms for probable PDD at Level I. The first set was based on the original framework [@dubois2007], which utilized the Mini-Mental State Examination (MMSE) as a global cognitive screening tool, supplemented by assessments of attention, executive function, visuospatial abilities, and memory. The second set of algorithms was based on the recent call for change of dementia diagnostic guidelines [@kulisevsky2024], which advocates for more sensitive cognitive domain assessments in the context of PD. This updated approach incorporated specific items from the MoCA. The third approach applied the Czech version of the shortened Montreal Cognitive Assessment (sMoCA) [@bezdicek2020], a time-efficient modification designed to measure global cognitive performance using a reduced testing protocol that omits items providing redundant information. **The sMoCA has been validated in the Czech PD cohort [@bezdicek2020] and shown to be sensitive to cognitive deficits while lowering patient burden [@roalf2017a]. We included the sMoCA in our study for its clinical utility in pre-surgical settings, where time restrictions and patients’ fatigue often limit the feasibility of longer assessments. Moreover, the Czech validation study reported comparable diagnostic accuracy between MoCA (AUC = 0.815) and sMoCA (AUC = 0.796) for distinguishing PD-MCI from PD-NC, supporting the sMoCA as a suitable and efficient alternative.**

@@

Following the framework proposed by @lundberg2021, in this study we explicitly connect our research objectives and their corresponding theoretical (i.e., targets of inference) and empirical (i.e., data-driven) estimands to statistical estimates. The theoretical estimand refers to a unit-specific quantity defined over a target population and represents the ideal quantity that would address the research question under optimal conditions, such as access to complete population data or perfect experimental control. In contrast, the empirical estimand corresponds to the quantity that is actually computable using the available dataset, given real world constraints. A full description of the study’s estimands and their relation to our research objectives is presented in the Appendix (see @tbl-estimands).

@@

# Tables of the 'best' and the 'worst' w.r.t. Level II:
approx_acc <- table_levelII_approximations(conc$table, algos, "Accuracy_raw")
approx_bal <- table_levelII_approximations(conc$table, algos, "Balanced Accuracy")

@@

**A total of** `r n_total` patients **were considered for the study, out of which `r excl_txt`**. The sample included `r sex` men, with an average of `r age`, `r edu`, `r dur`, `r updrs_off` and `r updrs_on`. **Descriptive statistics for neuropsychiatric symptoms indicated within average level of depressive and anxiety symptoms in our cohort, with the average BDI-II of `r bdi`, average STAI X1 of `r staix1`, and average STAI X2 of `r staix2`. However, according to the psychiatric assessment, none of the patients with probable PDD was suffering from the major depressive disorder, delirium or other neuropsychiatric abnormalities that would exclude the diagnosis.** Cognitive characteristics of the sample are summarized in @tbl-desc.

@@

```{r}
#| label: "fig-tree"
#| fig-cap: "A dendrogram representing algorithms for probable Parkinson's Disease Dementia (PDD) construction process."
#| apa-note: "The dendrogram illustrates the decision process used to construct algorithms for probable Parkinson’s Disease Dementia (PDD). The second level depicts the definition of instrumental activities of daily living (IADL) deficit (FAQ total > 7 on the left, FAQ item 9 > 1 on the right). The third level indicates the selection of the screening instrument (MMSE, MoCA, sMoCA, or none in the case of Level II). Lower branches represent the selection of neuropsychological tests used to define cognitive impairment in executive function, attention, memory, and language, ordered from top to bottom as depicted in the dendrogram. Algorithms based on the MMSE are shown in green, those based on the MoCA in blue, on the sMoCA in purple, and Level II algorithms in red. The top five screening algorithms according to raw accuracy for predicting diagnosis at Level II (see Table 3) are marked by bold edges. Accompanying dots indicate the estimated PDD rate for each algorithm. The individual test items comprising each algorithm are listed in Table A1."

make_dendrogram(rates$table, approx_acc$table)
```

@@

Results of the analyses of prediction Accuracy, Cohen's $\kappa$, Sensitivity and Specificity are presented in @fig-acc, @fig-kappa, @fig-sens and @fig-spec respectively. **The number of complete pairwise cases ranged from `r obs_pairs[1, "N"]` (`r min_miss` missing) to `r obs_pairs[nrow(obs_pairs), "N"]` (full sample). Most comparisons (`r perc_above_k`%) were based on `r K` or more observations.**[^1]

@@

approx_acc$gtable
```

@@

Our overall PDD rates were consistently lower than previous studies regarding PDD among PD patients, demonstrating wide variability based on various criteria used. For instance, a retrospective study reported a PDD rate of 19.7% [@rana2012], while other clinical investigations found an even rate higher rate, reaching up to 30% [@aarsland2005]. A recent meta-analysis synthesizing global data placed the expected PDD rate in PD at 26.30% [@sousa2022]. Compared to these estimates, our study reports generally lower PDD rates, likely reflecting differences in diagnostic criteria, methodology and sample characteristics. Specifically, our sample was younger compared to other PD cohorts and age was repeatedly shown to be a strong predictor of PDD across studies [@rana2012; @sousa2022].

@@

Furthermore, in the algorithm using sMoCA, the raw accuracy was moderate, however, the balanced accuracy (i.e. combined sensitivity and specificity) was high. **Consequently, sMoCA appears particularly suitable for approximating Level II PDD diagnosis in populations that differ in PDD prevalence from our cohort, since balanced accuracy, unlike raw accuracy, is independent of prevalence in the sample. Moreover, because the sMoCA algorithm demonstrated higher sensitivity while maintaining comparable specificity (see @tbl-approx-bal), it may be especially valuable in contexts where false negatives carry substantial clinical cost. In such cases, a neuropsychologist might use sMoCA as an initial screening tool and proceed to a full Level II assessment only for patients who meet criteria for probable PDD in this preliminary stage.**

@@

**Modern alternatives, such as multiple imputation techniques, have been shown to produce less biased estimates than case deletion strategies in analyses based on confusion matrices [@karakaya2014], whereas the evidence supporting their superiority in the estimation of Cohen’s $\kappa$ remains limited [@deraadt2019]. These modern approaches also require careful specification of the imputation model and the underlying causal mechanism of missingness to ensure appropriate covariate selection [@long2011; @bianco2023]. For the sake of parsimony, we did not employ advanced missing-data techniques but instead explicitly described the observed missingness patterns. Consequently, our findings should be regarded as exploratory and primarily serve as a basis for hypothesis generation in future, more controlled studies.**

@@

**An additional limitation concerns the use of the FAQ questionnaire for IADL assessment. The FAQ is a subjective or informant-reliant measure and thus susceptible to bias. Moreover, its content can vary across sociocultural contexts, limiting cross-cultural transferability. For example, activities such as financial management, cooking or driving, are not universally practised across societies. Consequently, both the FAQ scores and diagnostic thresholds used to indicate IADL impairment may not be directly transferable between cultural settings [@odonald2025]. These factors may influence both the sensitivity and ecological validity of the functional criteria used [@bezdicek2011; @bezdicek2016b].**

**Furthermore, IADL measures may correlate with neuropsychological results of specific cognitive domains, particularly attention/processing speed, and executive function [@reppermund2011; @moheb2017]. If such correlations were due to shared error variance, they could bias concordance indices measured in this study. To examine this possibility, we conducted a post-hoc simulation experiment, available at <https://josefmana.github.io/demcrit/articles/correlation.html>. The results indicated that correlations between IADL and neuropsychological measures may either increase or decrease accuracy and balanced accuracy, while consistently inflating Cohen’s $\kappa$ estimates. However, given the correlations observed in our dataset, the effect size was small and unlikely to alter the conclusions of our study.**

**To address concerns about measuring IADL outlined above, future research should consider using PD-specific questionnaires or more objective tools. Promising options include the Penn Parkinson's Daily Activities Questionnaire-15 [@brennan2016], questionnaire adaptations including items regarding gadgets use and digital literacy [@postema2024] or performance-based assessments [@schmitter-edgecombe2020]. Our findings underscore the importance of IADL measurement for the PDD diagnosis. Therefore, we recommend exploring more reliable tools with high ecological validity.**

Finally, whereas our study systematically investigated how varying definition of global deficit, impaired cognition and IADL deficit affect probable PDD classification, it did not explore associations of PDD diagnosis with its neuropsychiatric (e.g. anxiety profile) and biomarker correlates. **Instead, we only ensured the absence of acute or severe psychiatric symptoms that would preclude a diagnosis of probable PDD, as verified through assessment by a trained neuropsychiatrist. However, both the current diagnostic criteria for PDD [@dubois2007] and recent proposals for their revision [@kulisevsky2024] emphasize the use of standardized psychometric instruments for assessing neuropsychiatric symptoms. Incorporating such measures could enhance both the efficiency and transparency of the diagnostic process. Future research should therefore investigate how integrating structured neuropsychiatric assessments and biomarker data may refine PDD diagnostic accuracy and improve clinical utility.**

@@

Future studies should aim to replicate our results on larger and different cohorts **with more heterogeneous sample, and to explore how varying operationalization of other components of probable PDD, namely psychiatric symptoms, affects PDD rates and concordance between diagnostic algorithms**. To make this process easier, the code used to generate our results is publicly available and easily applicable to similarly structured data.

@@

In this study, we follow the framework proposed by @lundberg2021 for specifying targets of inference (i.e., the estimands) in quantitative sciences to increase transparency and connect statistical evidence to relevant theory. @tbl-estimands contains verbal description of the components relating to each of our proclaimed research objectives and map them to the population quantity of interest (the theoretical estimand), data-dependent quantity that could be estimated (the empirical estimand) and quantities that are reported in the study (statistical estimates).

@@

```{r}
#| label: "fig-rates"
#| fig-cap: "Summary of the estimates of probable PDD rate."
#| apa-note: "Vertical lines represent estimates arrived at by using sMoCA (dotted) or Level II (dashed) with FAQ item 9 (orange) or FAQ total score (blue) as criteria for probable PDD. The percentages were calculated from all available cases (see Table A3 for numerical summary)."

@@

approx_bal$gtable
```
